Using Fundamental Measure Theory to Treat the Correlation Function of the Inhomogeneous Hard-Sphere Fluid

### 2012-08-21 papers/contact/paper.tex
added first attempt at results writing Tested everything on node04.

Yu and Wu developed a functional for the correlation function
evaluated at contac, which should correspond to the symmetrically
averaged correlation
function~$g_\sigma^S$~\cite{yu2002fmt-dft-inhomogeneous-associating}.
Although Yu and Wu use a symmetric formulation for $g_\sigma$, instead
of using $n_0$ as the associated density, they use a density given by
\begin{align}
  n_\text{Yu}(\rr) &= n_0(\rr) \zeta(\rr) \\
  \zeta &= 1 - \frac{\mathbf{n_2}\cdot\mathbf{n_2}}{n_2^2}
\end{align}
where the function $\zeta$ is a measure of local inhomogeneity at the
point of contact.  Given this definition of density, their $g_\sigma$
is given by:
% \fixme{Which of the following two equations is more clear to you? We
%  only want one.}
\begin{align}
  %% g_\sigma^\text{Yu} &= \frac{1}{1-n_3}
  %%   + \frac32 \frac{\sigma n_2}{6}\frac{\zeta}{(1-n_3)^2}
  %%   + \frac12 \frac{\left(\frac{\sigma n_2}{6}\right)^2 \zeta}{(1-n_3)^3}
  %%   \\
  g_\sigma^\text{Yu} &= \frac{1}{1-n_3}
    + \frac14 \frac{\sigma n_2\zeta}{(1-n_3)^2}
    + \frac1{72} \frac{\sigma^2 n_2^2 \zeta}{(1-n_3)^3}
\end{align}
From this function, we extract a symmetric correlation function
\begin{align}
  g_\sigma^\text{Yu,S} &= \zeta^2\left(\frac{1}{1-n_3}
    + \frac14 \frac{\sigma n_2\zeta}{(1-n_3)^2}
    + \frac1{72} \frac{\sigma^2 n_2^2 \zeta}{(1-n_3)^3}\right)
\end{align}
which corresponds to our $g_\sigma^S$ and can be directly compared
with it.

@@

@@

@@

The derivatation of the asymmetrically averaged correlation function
$g_\sigma^A(\rr)$ requires that we express the free energy in terms of
a set of weighted densities in which the radius of the weighting
kernel is determined according to the position of the sphere in
question---which is to say, the location where the density itself is
evaluated.  This is the most natural way to construct such a
functional, in which the radius of spheres is a function of position.
Thus we define $n_3^A$ as
\begin{align}
  n_3^{A}(\rr') &= \int n(\rr'') \Theta(\left|\rr' - \rr''\right| -\sigma(\rr'')/2) d\rr''
\end{align}
and we define the other weighted densities similarly.  In order to
compute derivatives of the free energy with respect to $R(\rr)$, we
need to find $g_\sigma^A$.  In the case of $n_3^A$, we need
\begin{align}
  \frac{\delta n_3^{A} (\rr')}{\delta \sigma(\rr)} &=
  \frac 12 \int n (\rr'') \delta(|\rr' - \rr''| - \sigma(\rr'')/2) \delta(\rr-\rr'') d\rr'' \\
   &= n (\rr) \delta(|\rr' - \rr| - \sigma(\rr)/2)
\end{align}
Similarly, for $n_2^A$, we find
\begin{align}
  n_2^{A}(\rr') &= \int n(\rr'') \delta(|\rr' - \rr''| - \sigma(\rr'')/2) d \rr''\\
  \frac{\delta n_2^{A}(\rr')}{\delta \sigma(\rr)} &= -\frac 12 n(\rr) \delta'(|\rr' - \rr| - \sigma(\rr)/2)\\
\end{align}
This is somewhat more challenging, since it involves an unfamiliar
function, the derivative the Dirac delta function $\delta'$.
%
The derivatives of the remaining scalar densities can be reduced to
sums of the terms above:
\begin{align}
  n_1(\rr') &= \int \mathbf{dr''} \frac{n(\rr'')}{2\pi \sigma(\rr'')}
  \delta(|\rr'-\rr''| - \sigma(\rr'')/2) \\
  \frac{\delta n_1(\rr')}{\delta \sigma(\rr)}
  &= -\frac{n(\rr)}{4\pi
    \sigma(\rr)}\delta'(|\rr'-\rr| - \sigma(\rr)/2)
  -
  \frac{n(\rr)}{2\pi
    \sigma(\rr)^2}\delta(|\rr'-\rr| - \sigma(\rr)/2)
\end{align}
and
\begin{align}
  n_0(\rr') &= \int \mathbf{dr''} \frac{n(\rr'')}{\pi \sigma(\rr'')^2}
  \delta(|\rr'-\rr''| - \sigma(\rr'')/2) \\
  \frac{\delta n_0(\rr')}{\delta \sigma(\rr)}
  &= -\frac{n(\rr)}{2\pi
    \sigma(\rr)^2}\delta'(|\rr'-\rr| - \sigma(\rr)/2)
  -
  2 \frac{n(\rr)}{\pi
    \sigma(\rr)^3}\delta(|\rr'-\rr| - \sigma(\rr)/2)
\end{align}
The vector-weighted densities have no additional occurrences of $R$,
their functional derivatives come out similarly:
\begin{align}
  \mathbf{n}_{V2}^{A}(\rr') &= \int n(\rr') \delta(|\rr' - \rr''| - \sigma(\rr'')/2)
    \frac{\rr'-\rr''}{|\rr'-\rr''|} d \rr''\\
  \frac{\delta \mathbf{n}_{V2}^{A}(\rr')}{\delta \sigma(\rr)} &= \frac 12 n(\rr) \delta'(|\rr' - \rr| - \sigma(\rr)/2)
    \frac{\rr-\rr'}{|\rr-\rr'|}\\
\end{align}
and analogously for $\mathbf{n}_{V1}^A$:
\begin{align}
  \mathbf{n}_{V1}^A(\rr') &= \int d\rr'' \frac{n(\rr'')}{2\pi \sigma(\rr'')}
  \delta(|\mathbf{r'}-\rr''| - \sigma(\rr'')/2) \frac{\rr'-\rr''}{|\rr'-\rr''|}\\
  \frac{\delta \mathbf{n}_{V1}^A(\rr')}{\delta \sigma(\rr)}
  &= \frac{n(\rr)}{4\pi
    \sigma(\rr)}\delta'(|\rr'-\rr| - \sigma(\rr)/2) \frac{\rr-\rr'}{|\rr-\rr'|}
  +
  \frac{n(\rr)}{2\pi
    \sigma(\rr)^2}\delta(|\rr'-\rr| - \sigma(\rr)/2) \frac{\rr-\rr'}{|\rr-\rr'|}
\end{align}
This completes our exploration of the partial derivatives of weighted
densities of the $A$ variety.

>>>>

Yu and Wu developed a functional for the correlation function
evaluated at contac, which should correspond to the symmetrically
averaged correlation
function~$g_\sigma^S$~\cite{yu2002fmt-dft-inhomogeneous-associating}.
Although Yu and Wu use a symmetric formulation for $g_\sigma$, instead
of using $n_0$ as the associated density, they use a density given by
\begin{align}
  n_\text{Yu}(\rr) &= n_0(\rr) \zeta(\rr) \\
  \zeta &= 1 - \frac{\mathbf{n_2}\cdot\mathbf{n_2}}{n_2^2}
\end{align}
where the function $\zeta$ is a measure of local inhomogeneity at the
point of contact.  Given this definition of density, their $g_\sigma$
is given by: \fixme{Which of the following two equations is more clear
  to you? We only want one.}
\begin{align}
  g_\sigma^\text{Yu} &= \frac{1}{1-n_3}
    + \frac32 \frac{\sigma n_2}{6}\frac{\zeta}{(1-n_3)^2}
    + \frac12 \frac{\left(\frac{\sigma n_2}{6}\right)^2 \zeta}{(1-n_3)^3}
    \\
  g_\sigma^\text{Yu} &= \frac{1}{1-n_3}
    + \frac14 \frac{\sigma n_2\zeta}{(1-n_3)^2}
    + \frac1{72} \frac{\sigma^2 n_2^2 \zeta}{(1-n_3)^3}
\end{align}
From this function, we extract a symmetric correlation function
\begin{align}
  g_\sigma^\text{Yu,S} &= \zeta^2\left(\frac{1}{1-n_3}
    + \frac14 \frac{\sigma n_2\zeta}{(1-n_3)^2}
    + \frac1{72} \frac{\sigma^2 n_2^2 \zeta}{(1-n_3)^3}\right)
\end{align}
which corresponds to our $g_\sigma^S$ and can be directly compared
with it.

@@

We can see that the White Bear DFT calulations for the densities at varying distances from the respective surfaces match up well with the monte-carlo simulated densities.  The densities slope up towards and reach a maximum at the surfaces as the diffusion cutoff in that direction causes hard spheres to collect. We also see that $n_0$ slopes sharply downward towards the different surfaces, beginning the trend at about one radius away.  Part of the average that is taken for spheres in these locations will be taken within a volume for which the sphere density is uniformily zero. A similar trend is followed by $n_A$ beginning at about two radii away.  A sphere that is in contact with another at any position must be two radii away from the center of the first.  Thus, within two radii of the surface, one enters into a region where some of the contact volume is past the zero density boundary.

The S case correlation functions do not match the monte carlo simulated correlation as well as the functionals of Yu and Wu.  For the walls and the sperical cavity, the monte carlo simulations show this function going to zero at these surfaces, roughly beginning its downward trend at about one sphere radius out.  This stands to reason since that because there is an increased density up against the wall, there is little chance that points within a radius of the wall and the center of those spheres will be a contact point between two spheres, since the hard spheres cannot overlap.

The A correlation White Bear calculations match up very well with monte-carlo simulation.  Although they seem to be both very close, the Gross functional deos not have the near perfect overlap with simulation that the White Bear calculation deos.

@@

Although the general density attributes look similar to those of the low density calculations as one moves away from the surfaces, the range of difference is altered.  The spikes at the surfaces reach a height of roughly five times the homogeneous average, whereas with the low densities the difference is closer to a factor of $0.5$.  With a higher density of spheres to pack in those that are up against the wall, it is more difficult for those spheres to ever find their way out and away.  We also notice that the downward trend in both $n_0$ and $n_A$ have similar characteristics but are over all much more flat.  (Don't see why!)

We once again see that in the S case, although neither functionals tested fit perfectly, Yu and Wu's functional is a closer fit than the White Bear's.  We can see in the monte carlo's oscillations over increased distance from the surfaces that there are layers of spheres that are formed.  If one picks a point in one of these layters, that point is more likely to be a contact point between spheres, whereas if one picks a sphere a little over half a radius away, the likelyhood is at a minimum.  There is then another maximum at roughly one radius away from the surface.  (This suggests close packed layers?)

As with the lower densities, the A case White Bear fits to the monte carlo simulations very well.  The Gross functional seems to be accurate until it falls off close to surfaces.

@@

The derivatation of the asymmetrically averaged correlation function
$g_\sigma^A(\rr)$ requires that we express the free energy in terms of
a set of weighted densities in which the radius of the weighting
kernel is determined according to the position of the sphere in
question---which is to say, the location where the density itself is
evaluated.  This is the most natural way to construct such a
functional, in which the radius of spheres is a function of position.
Thus we define $n_3^A$ as
\begin{align}
  n_3^{A}(\rr') &= \int n(\rr'') \Theta(\left|\rr' - \rr''\right| -\sigma(\rr'')/2) d\rr''
\end{align}
and we define the other weighted densities similarly.  In order to
compute derivatives of the free energy with respect to $R(\rr)$, we
need to find $g_\sigma^A$.  In the case of $n_3^A$, we need
\begin{align}
  \frac{\delta n_3^{A} (\rr')}{\delta \sigma(\rr)} &=
  \frac 12 \int n (\rr'') \delta(|\rr' - \rr''| - \sigma(\rr'')/2) \delta(\rr-\rr'') d\rr'' \\
   &= n (\rr) \delta(|\rr' - \rr| - \sigma(\rr)/2)
\end{align}
Similarly, for $n_2^A$, we find
\begin{align}
  n_2^{A}(\rr') &= \int n(\rr'') \delta(|\rr' - \rr''| - \sigma(\rr'')/2) d \rr''\\
  \frac{\delta n_2^{A}(\rr')}{\delta \sigma(\rr)} &= -\frac 12 n(\rr) \delta'(|\rr' - \rr| - \sigma(\rr)/2)\\
\end{align}
This is somewhat more challenging, since it involves an unfamiliar
function, the derivative the Dirac delta function $\delta'$.
%
The derivatives of the remaining scalar densities can be reduced to
sums of the terms above:
\begin{align}
  n_1(\rr') &= \int \mathbf{dr''} \frac{n(\rr'')}{2\pi \sigma(\rr'')}
  \delta(|\rr'-\rr''| - \sigma(\rr'')/2) \\
  \frac{\delta n_1(\rr')}{\delta \sigma(\rr)}
  &= -\frac{n(\rr)}{4\pi
    \sigma(\rr)}\delta'(|\rr'-\rr| - \sigma(\rr)/2)
  -
  \frac{n(\rr)}{2\pi
    \sigma(\rr)^2}\delta(|\rr'-\rr| - \sigma(\rr)/2)
\end{align}
and
\begin{align}
  n_0(\rr') &= \int \mathbf{dr''} \frac{n(\rr'')}{\pi \sigma(\rr'')^2}
  \delta(|\rr'-\rr''| - \sigma(\rr'')/2) \\
  \frac{\delta n_0(\rr')}{\delta \sigma(\rr)}
  &= -\frac{n(\rr)}{2\pi
    \sigma(\rr)^2}\delta'(|\rr'-\rr| - \sigma(\rr)/2)
  -
  2 \frac{n(\rr)}{\pi
    \sigma(\rr)^3}\delta(|\rr'-\rr| - \sigma(\rr)/2)
\end{align}
The vector-weighted densities have no additional occurrences of $R$,
their functional derivatives come out similarly:
\begin{align}
  \mathbf{n}_{V2}^{A}(\rr') &= \int n(\rr') \delta(|\rr' - \rr''| - \sigma(\rr'')/2)
    \frac{\rr'-\rr''}{|\rr'-\rr''|} d \rr''\\
  \frac{\delta \mathbf{n}_{V2}^{A}(\rr')}{\delta \sigma(\rr)} &= \frac 12 n(\rr) \delta'(|\rr' - \rr| - \sigma(\rr)/2)
    \frac{\rr-\rr'}{|\rr-\rr'|}\\
\end{align}
and analogously for $\mathbf{n}_{V1}^A$:
\begin{align}
  \mathbf{n}_{V1}^A(\rr') &= \int d\rr'' \frac{n(\rr'')}{2\pi \sigma(\rr'')}
  \delta(|\mathbf{r'}-\rr''| - \sigma(\rr'')/2) \frac{\rr'-\rr''}{|\rr'-\rr''|}\\
  \frac{\delta \mathbf{n}_{V1}^A(\rr')}{\delta \sigma(\rr)}
  &= \frac{n(\rr)}{4\pi
    \sigma(\rr)}\delta'(|\rr'-\rr| - \sigma(\rr)/2) \frac{\rr-\rr'}{|\rr-\rr'|}
  +
  \frac{n(\rr)}{2\pi
    \sigma(\rr)^2}\delta(|\rr'-\rr| - \sigma(\rr)/2) \frac{\rr-\rr'}{|\rr-\rr'|}
\end{align}
\fixme{double-check the directions of all the vector quantities...}

This completes our exploration of the partial derivatives of weighted
densities of the $A$ variety.
